---
title: "🧬 Spatial Transcriptomics Analysis (part 2)"
format: html
---

# Welcome to our training on spatial transcriptomics data analysis

Spatial transcriptomics allows us to measure gene expression across tissue sections while preserving spatial information, helping us understand not just **what** genes are active, but **where** they are active.

In this practical session, we will explore how to analyze spatial transcriptomics data using **R** and **RStudio**. Whether you are new to spatial data or already familiar with single-cell workflows, this guide will walk you through the essential steps.

Feel free to use your personal workbook to make notes, track your progress, and revisit it after this workshop.

---

In this **second part**, we will cover key steps in the analysis workflow, including quality control, preprocessing, clustering and visualization, cluster annotation, spatial clustering, and integration with single-cell transcriptomics data.

## Setting up your R environment 

Before diving into the data, we need to load a few important R packages. These libraries will provide the tools we need for analysis and visualization.

```{r}
#| label: setup
#| message: true

library(Seurat)     # Main package for single-cell and spatial transcriptomics analysis
library(SeuratWrappers) # Add-ons for Seurat
library(patchwork)  # Helps comcelle multiple ggplot2 plots into one figure
library(tidyverse)  # Essential collection of R packages, including ggplot2, dplyr, magrittr
library(grid)       # For other plots
library(viridis)    # Our continuous color palette

# Colors
cal_pal50 = c("#Fa1a8e", "#009B7D", "#ff9933", "#7083a4", "#ffce45", "#015e05", 
              "#fedf25", "#d2b48c", "#bb55e1", "#6ec5ab", "#5d57af", "#143341", 
              "#761445", "#d65b5a", "#94043a", "#e7a6cd", "#204519", "#87afd1", 
              "#9b9a9d", "#f95b70", "#83c874", "#808080", "#452b5b", "#ecb100", 
              "#f46124", "#525252", "#4c84a3", "#00bfff", "#01b4c6", "#174d79", 
              "#a6a0f2", "#76facc", "#8491be", "#a32a2f", "#1c8859", "#2cc012", 
              "#35782b", "#9c6755", "#3b3960", "#eeb0a1", "#3e1e13", "#0064c3", 
              "#d81e4a", "#74646c", "#f675da", "#ffce45", "#ec7014", "#e50000", 
              "#000000", "#a4527c", "#041859")

# Set default colours for ggplot2
options(ggplot2.discrete.colour=cal_pal50)
options(ggplot2.discrete.fill=cal_pal50)
options(ggplot2.continuous.colour="viridis")
options(ggplot2.continuous.fill="viridis")
options(ggplot2.cellned.colour="viridis")
options(ggplot2.cellned.fill="viridis")
```

# Loading the dataset

```{r}
#| label: dataset
# Load 10x Xenium dataset with Seurat
sc = LoadXenium("datasets/xenium_mouse_brain", fov="fov", segmentations="cell")

sc$orig.ident = "brain"
Idents(sc) = "brain"

# Load also nucleus_area and cell_area from cells.csv.gz
cells_metadata = read.csv("datasets/xenium_mouse_brain/cells.csv.gz")
sc = AddMetaData(sc, metadata=cells_metadata %>% select(cell_id, nucleus_area, cell_area) %>% column_to_rownames("cell_id"))
sc$nucleus_area = ifelse(is.na(sc$nucleus_area), 0, sc$nucleus_area) # there is a bug in Seurat ImageFeaturePlot if NA is in the data
```

# Preface

The Xenium platform provides spatially resolved transcriptomics readouts. For each and every transcript, we know the original physical coordinates within the tissue. This means that transcripts can be analyzed in their spatial context, even before defining individual cell boundaries. Such analyses are referred to as **segmentation-free analyses**.

Alternatively, we can delineate cell boundaries with **cell segmentation** methods. Thinking in terms of cells often simplifies data interpretation and allows us to use the extensive toolkit developed for scRNA-seq analysis. However, these scRNA-seq tools do not incorporate spatial information, which limits their ability to fully exploit the richness of Xenium data. Consequently, dedicated methods for spatially resolved transcriptomics are actively being developed.

In this workflow, we work with cell segmentation. The data we read and analyse are classical gene-by-cell count matrices. We begin with standard quality control (QC) and preprocessing steps commonly used in scRNA-seq analysis. This provides an initial understanding of the data before applying spatially aware methods.

# Quality control

## Why quality control?

In spatial transcriptomics (and any sequencing-based assay), not all cells are equally informative. Quality Control (QC) is used to identify and remove low-quality cells from the dataset before further analysis. 

**Low-quality cells** might be the result of:  

* Problems during reverse transcription, PCR amplification or library preparation  
* Cell damage during the experiment, e.g. when the tissue already contains a lot of dead or dying cells  
* Very few transcripts overall, e.g. in empty regions  
* Unusually high numbers of transcripts, possibly doublets or noise  

**Why does this matter?**  
Low-quality cells can distort the results. For instance, they might cluster separately and be falsely interpreted as novel biological states or cell types. Performing QC early helps ensure the reliability of downstream steps like normalization, clustering, and cell type annotation.

## QC metrics at a glance

We will use three common metrics to flag low-quality cells:

- **Number of counts (UMIs):** Measures total expression in a cell; too low = empty; too high = potential artifacts.
- **Number of detected genes:** Captures the diversity of expression in a cell; low gene counts might mean poor capture.
- **Mitochondrial gene content:** High levels often indicate stressed or dying cells.

These metrics will be assessed both statistically and spatially to get an idea of the sample quality and to identify thresholds for filtering.

## Number of counts per cell

The total number of UMIs (counts) per cell gives a sense of overall transcript abundance.

```{r}
#| label: qc_1
#| warning: false

p1 = VlnPlot(sc, features="nCount_Xenium", pt.size=0, layer="counts", adjust=.5) + NoLegend()
p2 = ImageFeaturePlot(sc, features="nCount_Xenium", max.cutoff="q95") + viridis::scale_fill_viridis()
p1 | p2

# Look at the summary statistics
summary(sc$nCount_Xenium)

# Find a cutoff
cut_nCount = 30
p3 = ggplot(sc[[]], aes(x=nCount_Xenium)) + geom_histogram(fill="black", bins=500) + xlim(c(0, 1000)) + theme_minimal() + 
  geom_vline(xintercept=cut_nCount, color=cal_pal50[1]) + 
  ggtitle("Histogram of nCount_Xenium")
p3
```

Look for cells with very few or very many counts. cells with <`cut_nCount` counts are suspicious and flagged for removal. 

::: {.callout-tip}
### Cells flagged here should appear randomly across the tissue #####
If they form a distinct pattern, reconsider the threshold — you might be removing biologically relevant data.
:::

```{r}
#| label: qc_2

sc$filtered_nCount = ifelse(sc$nCount_Xenium<cut_nCount, "filtered", "kept")
ImageDimPlot(sc, group.by="filtered_nCount", cols=c(filtered="#FFFF00", kept="grey50"))
```

::: {.callout-note}
### Find your optimal threshold
Play around with the thresholds in R. Which threshold would you find best?
:::

## Number of genes per cell

This tells us how many genes are detected per cell — essentially a measure of complexity.

```{r}
#| label: qc_3

p1 = VlnPlot(sc, features="nFeature_Xenium", pt.size=0, layer="counts", adjust=.5) + NoLegend()
p2 = ImageFeaturePlot(sc, features="nFeature_Xenium", max.cutoff="q95") + viridis::scale_fill_viridis()
p1 | p2

# Look at the summary statistics
summary(sc$nFeature_Xenium)

# Find a cutoff
cut_nFeature = 15
p3 = ggplot(sc[[]], aes(x=nFeature_Xenium)) + geom_histogram(fill="black", bins=50) + theme_minimal() +
  geom_vline(xintercept=cut_nFeature, color=cal_pal50[1]) + 
  ggtitle("Histogram of nFeature_Xenium")
p3
```

```{r}
#| label: qc_4

sc$filtered_nFeature = ifelse(sc$nFeature_Xenium<cut_nFeature, "filtered", "kept")
ImageDimPlot(sc, group.by="filtered_nFeature", cols=c(filtered="#FFFF00", kept="grey50"))
```

## Mitochondrial content

Mitochondrial RNA (mtRNA) often increases when cells are stressed or dying. A high percentage is a red flag.

**If the Xenium panel includes mitochondrial genes**, we identify these genes using a prefix ("^mt-" means all genes starting with "mt-") and calculate their percentage per cell:

```{r}
#| label: qc_5
#| eval: false # only apply this chunk if mitochondrial genes are included in the dataset

#sc = PercentageFeatureSet(sc, pattern="^mt-", col.name="pMito_Xenium") 

#p1 = VlnPlot(sc, features="pMito_Xenium", pt.size=0, layer="counts") + NoLegend()
#p2 = SpatialFeaturePlot(sc, features="pMito_Xenium") + NoLegend()
#p1 | p2

# Look at the summary statistics
#summary(sc$pMito_Xenium)
```

## Cell area

With cell segmentation available, we can also inspect cell size metrics like `cell_area` and `nucleus_area`. Extremely small or large cells might indicate segmentation errors or artifacts.

```{r}
#| label: qc_6

p1 = ImageFeaturePlot(sc, features="cell_area", max.cutoff="q95") + viridis::scale_fill_viridis()
p2 = ImageFeaturePlot(sc, features="nucleus_area", max.cutoff="q95") + viridis::scale_fill_viridis()
p1 | p2
```

## Cell segmentation method 

Now that we got an impression on the QC metrics for the dataset, let us have a look at how cells were segmented in the first place, and how the segmentation relates to the QC metrics: 

```{r}
#| label: qc_6_2

# Number of cells in the dataset
n_cells = length(Cells(sc))

# Cell segmentation method percentages
table(sc$segmentation_method)
table(sc$segmentation_method) %>% prop.table() * 100

# Visualize segmentation methods
cols = cal_pal50[c(20, 40, 50)]
names(cols) = unique(sc$segmentation_method)
p1 = ImageDimPlot(sc, group.by="segmentation_method", cols=cols) + ggtitle(paste(n_cells, "cells total"))
p1

# How does the segmentation method relate to nCount?
p2 = VlnPlot(sc, features="nCount_Xenium", group.by="segmentation_method", pt.size=0, cols=cols, adjust=.5, log=TRUE) + 
  theme(axis.text.x=element_blank(), legend.position="bottom") + 
  guides(fill=guide_legend(ncol=1)) +
  geom_hline(yintercept=cut_nCount, color=cal_pal50[1])

# How does the segmentation method relate to nFeature?
p3 = VlnPlot(sc, features="nFeature_Xenium", group.by="segmentation_method", pt.size=0, cols=cols, adjust=.5, log=TRUE) + 
  theme(axis.text.x=element_blank(), legend.position="bottom") + 
  guides(fill=guide_legend(ncol=1)) +
  geom_hline(yintercept=cut_nFeature, color=cal_pal50[1])
(p2 | p3) + plot_layout(guides = "collect") & theme(legend.position="bottom")

# Repeat the same violin plot with cell_area and nucleus_area
p4 = VlnPlot(sc, features="cell_area", group.by="segmentation_method", pt.size=0, cols=cols, adjust=.5, log=TRUE) + 
  theme(axis.text.x=element_blank(), legend.position="bottom") + 
  guides(fill=guide_legend(ncol=1))

p5 = VlnPlot(sc, features="nucleus_area", group.by="segmentation_method", pt.size=0, cols=cols, adjust=.5) + 
  theme(axis.text.x=element_blank(), legend.position="bottom") + 
  guides(fill=guide_legend(ncol=1))
(p4 | p5) + plot_layout(guides = "collect") & theme(legend.position="bottom")

# Plot the cells with no nucleus_area
sc$nucleus = ifelse(sc$nucleus_area==0, "no_nucleus", "has_nucleus")
cols_nucleus = c(no_nucleus="#FFFF00", has_nucleus="grey50")
p6 = ImageDimPlot(sc, group.by="nucleus", cols=cols_nucleus) + ggtitle(paste(sum(sc$nucleus=="no_nucleus")), "cells without nucleus")
p6
```

::: {.callout-note}
### Questions  
What do you think about the relation of low QC and segmentation by nuclear expansion?
Why might mitochondrial content be misleading in some tissues (e.g., muscle or brain)?  
What are the trade-offs of stricter vs. more lenient filtering?  
:::

## Filtering of cells

Once you have reviewed the metrics and decided on thresholds, remove poor-quality cells using subset():

Before filtering:

```{r}
#| label: qc_7

sc
```

Apply the filters:

```{r}
#| label: qc_8

sc = subset(sc, nCount_Xenium < cut_nCount, invert=TRUE) %>% suppressWarnings()
sc = subset(sc, nFeature_Xenium < cut_nFeature, invert=TRUE) %>% suppressWarnings()
# sc = subset(sc, pMito_Xenium > cut_pMito, invert=TRUE) %>% suppressWarnings()
sc
```

## Filtering of genes

Genes expressed in only a few cells can add noise. First, we count how many cells each gene is detected in:

```{r}
#| label: qc_9

counts = GetAssayData(sc, layer="counts")
genes_expressed = rowSums(counts > 0)
head(genes_expressed, 5)
```

Let us quickly summarize these numbers: 

```{r}
#| label: qc_10

summary(genes_expressed)
```

If required, keep genes expressed in >N cells:

```{r}
#| label: qc_11
#| eval: false

#genes_expressed = genes_expressed[genes_expressed > 10]
#length(genes_expressed)
```

Then subset the Seurat object:

```{r}
#| label: qc_12
#| eval: false

#sc
#sc = subset(sc, features=names(genes_expressed)) %>% suppressWarnings()
#sc
```

::: {.callout-tip}
### Best practices
* Avoid "one-size-fits-all" thresholds; adjust for every sample
* Make sure to check the spatial distribution of filtered cells to avoid removing spatial patterns
* At the beginning, run a full analysis (until clustering) without filtering 
  * Low-quality cells often cluster together
  * Plot QC metrics per cluster and identify clusters without or with ambiguous cell type
  * Then adjust filtering criteria accordingly and repeat
:::

# Preprocessing

After removing low-quality cells, we now prepare the dataset for downstream analysis. This **preprocessing pipeline** includes:

1. Normalization of raw counts
2. Identification of highly variable genes (HVGs)
3. Centering and scaling of those HVGs
4. Optional sketching of the dataset for speed (HVGs)
5. Dimensionality reduction (PCA) based on the HVGs

Each step ensures that only informative, reliable data is passed into clustering and annotation workflows.

## Normalization

Raw counts vary widely across cells due to differences in sequencing depth (library size). To make counts **comparable across cells**, we perform normalization.

Note that normalization methods for spatial data are still being developed and evaluated. For this workshop, we apply log-normalization. 

**How does it work?**  

* Per cell, the count for each gene is divided by the total count for that cell, and then multiplied by 10,000  
* These "counts per 10,000" are then log-transformed using the natural logarithm (with a pseudocount of 1)  

This is implemented in Seurat with `NormalizeData()`:

```{r}
#| label: preprocessing_option_1
#| eval: false

# Load a function that we wrote to copy raw "counts" to the "data" layer without normalization
#source("scripts/additionalFunctions.R") # load TransformData function

#sc = TransformData(sc, assay="Xenium", layer="counts", save="data", log=TRUE)
```

```{r}
#| label: preprocessing_option_2
#| eval: false

# sc = NormalizeData(sc, normalization.method="LogNormalize", scale.factor=100)
```

```{r}
#| label: preprocessing_option_3
#| eval: true

source("scripts/additionalFunctions.R") # load SpaNormWrapper function
sc = SpaNormWrapper(sc=sc)
saveRDS(sc, file="sc_normalized.RDS")
```

This creates a new layer in the assay object called `data` that contains the normalized counts. You can inspect the normalized values like so:

```{r}
#| label: preprocessing_2

sc = readRDS("sc_normalized.RDS")

data = GetAssayData(sc, layer="data")
data[1:10, 1:10]
```

## Identifying Highly Variable Genes (HVGs)

Not all genes are equally informative. Experience shows that 1,000-3,000 genes with the highest cell-to-cell variation are often sufficient to describe the global structure of a dataset. 

Highly variable genes are typically the genes that show strong variation across cells and are more likely to distinguish biological states. Housekeeping genes with similar expression profiles across all cells, or genes with minor expression differences, might add random noise and mask relevant changes during downstream analysis. 

Here, we determine the top HGVs:

```{r}
#| label: preprocessing_3

sc = FindVariableFeatures(sc)
```

Get the top 10 HVGs and plot them:

```{r}
#| label: preprocessing_4

top10 = VariableFeatures(sc) %>% head(10)
top10

p = VariableFeaturePlot(sc, selection.method="vst")
p = LabelPoints(plot=p, points=top10, repel=TRUE)
p
```

## Data scaling

Gene expression ranges vary; some genes have inherently higher values. Centering and scaling ensures each gene contributes equally during the analysis. By centering (mean expression of 0) and scaling (standard deviation of 1), all genes will be on the same scale which makes them comparable. 

We use the `ScaleData` function to center and scale the HVGs:

```{r}
#| label: preprocessing_5

sc = ScaleData(sc, features=VariableFeatures(sc))
GetAssay(sc, assay="Xenium")
```

This creates a new layer in the assay object called `scale.data` that contains the scaled counts:

```{r}
#| label: preprocessing_6

scale_data = GetAssayData(sc, layer="scale.data")
scale_data[1:5, 1:5]
```

## Dimensionality reduction

With ~20,000 genes, the data is too high-dimensional for visualization or clustering. At this point of the analysis, we have already reduced the dimensionality of the dataset to HVGs. The biological manifold however can be described by far fewer dimensions than the number of (variable) genes, since expression profiles of different genes are correlated if they are involved in the same biological process. 

We use Principal Component Analysis (PCA) to reduce the dimensionality. PCA finds new axes (PCs) that capture the most variance in the dataset, and projects the data onto these axes. The resulting meta-variables are called principal components (PCs).

We use the `RunPCA` function to calculate the top 50 principle components (PCs) for the data. This creates a new dimensionality reduction object called `pca`:

```{r}
#| label: preprocessing_10

sc = RunPCA(sc, reduction.name="pca", features=VariableFeatures(sc), npcs=50, nfeatures.print=5)
Reductions(sc, slot="pca")
```

Visualize first two PCs:

```{r}
#| label: preprocessing_11

DimPlot(sc, reduction="pca", dims=c(1, 2))
```

::: {.callout-note}
## Assignment

Visualize the PCA for PC2 and PC3. 
:::

::: {.callout-note}
### Question
What do we learn from the PCA plot? 
:::

In PCA, genes with the largest variation between cells will have the biggest influence on the PCs. The first PC captures the greatest variance across the dataset. The next PC captures the greatest remaining amount of variance, and so on. This way, the top PCs are likely to represent the biological signal where multiple genes are affected by the same biological processes in a coordinated way. In contrast, random technical or biological noise that affects each gene independently are contained in later PCs. 

Downstream analysis can be restricted to the top PCs. We need to decide how many PCs we want to use for our analysis. The following elbow plot shows PCs ranked based on the variance they explain. 

Look for the "elbow" — a point where adding more PCs gives diminishing returns:

```{r}
#| label: preprocessing_12

ElbowPlot(sc, reduction="pca", ndims=50)
```

You can also check which genes contribute most to each PC:

```{r}
#| label: preprocessing_13

VizDimLoadings(sc, reduction="pca", dims=1:4, nfeatures=10, balanced=TRUE)
```

::: {.callout-tip}
## Learnings
Helpful YouTube video on PCA in the context of gene expression analysis is [here](https://www.youtube.com/watch?v=FgakZw6K1QQ)
:::

::: {.callout-note}
### Question
How many PCs would you choose for further analysis? 
:::

# Clustering and visualization

In the following, we **cluster** spatial cells based on their gene expression profiles and use **UMAP** (Uniform Manifold Approximation and Projection) to visualize the results in two dimensions.

**Clustering** is typically done using unsupervised algorithms (e.g., Louvain or Leiden) that group cells based on similarities across the full gene expression space. This helps to identify regions within the tissue that share similar transcriptional signatures, which may correspond to distinct anatomical structures or functional states. 

**UMAP** is a dimensionality reduction technique that projects high-dimensional data into a lower-dimensional space, in this case 2D, to reveal underlying structure. In UMAP plots, each cell is shown as a point, allowing us to detect substructures or transitions within clusters, identify outliers or ambiguous regions, communicate findings visually and intuitively. 

Clusters are discrete groupings assigned algorithmically based on expression similarity; each cell belongs to exactly one cluster. UMAP clouds, in contrast, are continuous visual projections. While clusters often appear as distinct clouds in UMAP space, overlaps can occur, especially in biologically complex areas. UMAP may reveal gradual transitions or relationships that clustering alone cannot capture.

## Clustering

Seurat uses a graph-based approach for clustering. In short, Seurat first constructs a graph by identifying the k-nearest neighbors for each cell in the PCA space based on Euclidean distance (NN). It then builds a refined version of the neighbor graph by considering shared neighbors between cells (SNN). Finally, a clustering algorithm is applied to the graph to identify groups of similar cells. These groups are then considered distinct cell populations. 

We begin by creating a nearest-neighbor graph using the `FindNeighbors` function in Seurat. We use the first 30 PCs as input and set the number of nearest neighbors to 20:

```{r}
#| label: clustering_1

sc = FindNeighbors(sc, reduction="pca", dims=1:30, k.param=20)
Graphs(sc)
```

After constructing the nearest-neighbor graph, we apply the Leiden algorithm to perform the clustering. The Leiden algorithm is an improvement over the Louvain method and works by optimizing modularity. It groups cells into clusters such that cells within the same cluster are more connected to each other than to cells in other clusters.

We call the `FindClusters` function with `algorithm="leiden"`, which tells Seurat to use the Leiden algorithm. We also need to specify the resolution parameter. This parameter controls the granularity of the clustering:

* Higher resolution values lead to more clusters  
* Lower resolution values lead to fewer clusters  

We will start with a resolution of 2, but this can be adjusted later depending on the number of clusters you wish to identify:

```{r}
#| label: clustering_2

# method="igraph" is necessary for Seurat versions < 5.2
sc = FindClusters(sc, resolution=.5, algorithm="leiden", random.seed=1)
```

The clusters are saved as cell metadata column `seurat_clusters`. Let us summarize the clustering of the dataset:

```{r}
#| label: clustering_3

table(sc$seurat_clusters)
```

Note that we can provide multiple resolution values to test different clustering resolutions.

::: {.callout-note}
## Assignment
Test different cluster resolutions and observe the number of clusters you receive. 
:::

```{r}
#| label: clustering_colors
cols_clusters = cal_pal50[1:length(levels(sc$seurat_clusters))]
names(cols_clusters) = levels(sc$seurat_clusters)
```

## UMAP 

Now, we move on to generating a UMAP plot, which provides a visual representation of the dataset in lower-dimensional space. This helps to visually identify groups of similar cells, making it easier to explore and interpret the data.

We run the `RunUMAP` function and use the first 30 PCs as input. This creates a new dimensionality reduction object called `umap`:

```{r}
#| label: clustering_4

sc = RunUMAP(sc, reduction="pca", reduction.name="umap", dims=1:30)
```

Note that `umap` is now the default dimensionality reduction. We can color the clustering on the UMAP as well as in spatial context:

```{r}
#| label: clustering_5

p1 = DimPlot(sc, reduction="umap", label=TRUE, cols=cols_clusters) + NoLegend()
p2 = ImageDimPlot(sc, group.by="seurat_clusters", cols=cols_clusters) + NoLegend()
p1 | p2
```

::: {.callout-note}
## Assignment

The parameter `dims` specifies how many PCs (and hence information) we use as input for UMAP. Similarly, the parameter `n.neighbors` controls how UMAP balances local (few neighbors) versus global structure (many neighbors) in the data.

Take some time to play around with these two parameters. What happens, if you only use the first 5 PCs?
Plot the old and new UMAP next to each other for comparison.
:::

```{r}
#| label: clustering_4_5_test
sc = RunUMAP(sc, reduction="pca", reduction.name="umap.test", dims=1:5, 
             n.neighbors=30)

p1 = DimPlot(sc, reduction="umap", label=TRUE) + NoLegend()
p2 = DimPlot(sc, reduction="umap.test", label=TRUE) + NoLegend()
p1 | p2
```

When dealing with many clusters, some of which may be spatially restricted while others are not, it can be challenging to interpret their spatial distribution. In such cases, highlighting individual clusters can provide more clarity.

We can highlight specific clusters by using the `CellsByIdentities` function to select cells belonging to particular clusters. Here, we highlight clusters 16 and 8:

```{r}
#| label: clustering_6

cols_cluster_cells = ifelse(levels(sc$seurat_clusters) %in% c(11), "#FFFF00", "grey50")
names(cols_cluster_cells) = levels(sc$seurat_clusters)

p1 = ImageDimPlot(sc, group.by="seurat_clusters", cols=cols_cluster_cells)
p1
```

::: {.callout-tip}
## Recommendations

* Clustering  
  * Test different resolution values to find the best clustering solution for your data  
* UMAP  
  * UMAP "cloud" sizes and the distance between them may not be meaningful since the method has a local notion of distance
  * Helpful explanation on YouTube [here](https://www.youtube.com/watch?v=eN0wFzBA4Sc)
  * Helpful documentation and option to explore [here](https://pair-code.github.io/understanding-umap/)  
* Take your time. This is perhaps the most time-consuming step of the basic analysis
:::

::: {.callout-note}
## Question   
* How many clusters do we have?    
* What can we already say about the clusters we have now?   
:::

## Spatial clustering

Up to now, clusters were driven only by gene expression. Here we add spatial context so that nearby cells are more likely to be grouped together when their expression is compatible. We follow the SpatialLeiden workflow (ported from the Python tutorial: https://spatialleiden.readthedocs.io/stable/usage.html) but run it from R via reticulate.

First, export the minimum needed from Seurat into a Python AnnData: normalized counts, PCA embedding, and cell centroids.

```{r}
#| label: spatial_clustering_1

# Get normalized counts and transpose (in python matrices are backwards)
data = SeuratObject::LayerData(sc, assay="Xenium", layer="data")
barcodes = colnames(data)
features = rownames(data)
data = Matrix::t(data)

# Get PCA
pca = SeuratObject::Embeddings(sc, "pca")
pca = as.matrix(pca)

# Get centroids of cells 
coords = SeuratObject::GetTissueCoordinates(sc[["fov"]])
rownames(coords) = coords$cell
coords$cell = NULL
coords = as.matrix(coords)

# Load python modules (add convert=FALSE to TRUE automatic conversion from python to R and back)
# - E.g. load scanpy module and name it "scanpy"
# - All python calls: Use "$" instead of ".". E.g. scanpy.AnnData => scanpy$AnnData
# - All basic data types are converted automatically. '{}' becomes list. '[]' becomes 'c()'. Sometimes you need to force a type though (as.integer).
# - Also check functions py_to_R and R_to_py
scanpy = reticulate::import("scanpy", convert=FALSE)
numpy = reticulate::import("numpy", convert=FALSE)
scipy = reticulate::import("scipy", convert=FALSE)
squidpy = reticulate::import("squidpy", convert=FALSE)
spatialleiden = reticulate::import("spatialleiden", convert=FALSE)

# Convert into minimal python anndata object
adata = scanpy$AnnData(X = scipy$sparse$csr_matrix(data),
                       obsm = list("spatial"=coords, "X_pca"=pca))
adata$var_names = features
adata$obs_names = barcodes
adata
```

Build two neighbor graphs: one in expression space (Scanpy) and one in physical space (Squidpy), then convert spatial distances to connectivity.

```{r}
#| label: spatial_clustering_2

# Get neighbors graph based on gene expression
scanpy$pp$neighbors(adata, random_state=as.integer(42))

# Get neighbors graph based on spatial context
squidpy$gr$spatial_neighbors(adata, coord_type="generic", n_neighs=as.integer(10))

# Calculate distances between neighbors
adata$obsp["spatial_connectivities"] = spatialleiden$distance2connectivity(
    adata$obsp["spatial_distances"]
)
```

Run SpatialLeiden to combine both graphs into a single clustering, controlling granularity with resolution and the spatial-vs-expression tradeoff with layer_ratio.

```{r}
#| label: spatial_clustering_3

# Spatial clustering
spatialleiden$spatialleiden(
    adata, resolution=2, layer_ratio=1.5, directed=c(FALSE, TRUE), random_state=as.integer(42)
)

# Translate results back into R
cluster_results = as.data.frame(reticulate::py_to_r(adata$obs))
sc = AddMetaData(sc, metadata=cluster_results)

# Let's focus on spatial clusters from here on 
Idents(sc) = "spatialleiden"
```

Bring the cluster labels back into Seurat metadata and compare tissue plots of expression-only vs spatially informed clusters.

```{r}
#| label: spatial_clustering_4

p1 = ImageDimPlot(sc, group.by="seurat_clusters", cols=cols_clusters) 
p2 = ImageDimPlot(sc, group.by="spatialleiden") 
p1 | p2
```

# Cluster annotation

What cell types correspond to the clusters? To identify the cell types corresponding to the clusters, we typically follow these steps:

* **Cluster marker genes**: Use differential gene expression analysis to find genes that are specifically expressed in each cluster. These can serve as markers for the corresponding cell types.
* **Known marker genes**: Visualize the expression of known cell-type markers on the UMAP to help associate clusters with specific cell types.
* **Mapping a reference dataset**: Use external datasets to map cell-type labels onto your data, enhancing the accuracy of your cell-type predictions.

By using these strategies, you can correlate the clusters with biological cell types and better understand the cell composition of your dataset.

## Cluster marker genes

One of the first steps in identifying the cell types of each cluster is to determine cluster-specific marker genes. These are genes that are significantly up- or down-regulated in one cluster compared to all other clusters, and they can be used to characterize the biological identity of the cells in each cluster. 

A good clustering of cells typically results in good marker genes. Hence, if you cannot find good marker genes you may need to go back to the start of the workflow and adapt your parameters. 

Resulting *p*-values are adjusted using the Bonferroni method. However, beware that *p*-values are likely inflated, since both clusters and marker genes were determined based on the same gene expression data, and there ought to be gene expression differences by design. Nevertheless, *p*-values can be used to sort and prioritize marker genes. 

You can use Seurat `FindMarkers` function to identify differentially expressed genes between clusters:

```{r}
#| label: clusterannot_1

markers = FindAllMarkers(object=sc,
                         test.use="wilcox",
                         layer="data",
                         only.pos=FALSE,
                         max.cells.per.ident=1000)
```

Let us now extract and visualize the top 3 marker genes for each cluster based on the adjusted p-value and log2 foldchange: 

```{r}
#| label: clusterannot_2

# Solution
markers_top = markers %>% 
  dplyr::group_by(cluster) %>% 
  dplyr::arrange(p_val_adj, avg_log2FC) %>% 
  dplyr::slice_head(n=3) %>% 
  dplyr::ungroup() %>% 
  as.data.frame()

gt::gt(markers_top) %>% 
  gt::tab_options(container.height=450)
```

</br>

Good marker genes are highly and possibly even only expressed in one cluster as compared to all other clusters. However, sometimes marker genes are also expressed in other clusters, or are declared as marker genes in these clusters, for example cell lineage markers that are shared by different cell sub types. To evaluate marker genes, it is essential to visualize their expression patterns.

```{r}
#| label: clusterannot_3
#| fig-height: 8
#| fig-width: 12

DotPlot(sc, features=markers_top$gene %>% unique()) +
  viridis::scale_color_viridis() + 
  ylab("Cluster") + xlab("") +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=.5),
        legend.position="bottom") + 
  guides(size=guide_legend(order=1, title="Pct expressed"), color=guide_colorbar(title="Scaled Expression")) + 
  ggtitle("Top markers per cluster (expression scaled)")

DotPlot(sc, features=markers_top$gene %>% unique(), scale=FALSE) + 
  viridis::scale_color_viridis() + 
  ylab("Cluster") + xlab("") + 
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=.5),
        legend.position="bottom") + 
  guides(size=guide_legend(order=1, title="Pct expressed"), color=guide_colorbar(title="Average Expression")) +
  ggtitle("Top markers per cluster (expression unscaled)")

```

::: {.callout-note}
## Assignment

Take a good look at the dot plot. Which clusters do / do not have good marker genes? 
:::

## Functional enrichment analysis

We can use the identified cluster marker genes to perform functional enrichment analysis. This helps us infer the potential biological identity or function of each cluster. For simplicity, we will use the online tool [Enrichr](https://maayanlab.cloud/Enrichr/) to perform this analysis. 

::: {.callout-note}
## Assignment

Try to find the identity of one cluster! 
:::

## Plotting known marker genes

To further explore the results and visualize gene expression for known marker genes across different clusters, we can use various plotting functions.

A **violin plot** allows us to visualize the distribution of gene expression across clusters. For example, we can plot the expression of genes Acta2 and Apoe:

```{r}
#| label: plotting_1

features_to_plot = c(Cortex1="Rorb", Cortex2="Satb2", Ventricle="Sox2", Midbrain="En1")

plist = VlnPlot(sc, features=features_to_plot, pt.size=0, ncol=1, combine=FALSE)
for (i in seq(plist)) { print(plist[[i]] + NoLegend()) }
```

We can also visualize the gene expression directly on top of the **UMAP plot**. This will show the expression of genes such as Acta2 and Apoe across all cells in the low-dimensional UMAP space:

```{r}
#| label: plotting_2

plist = FeaturePlot(sc, features=features_to_plot, ncol=2, max.cutoff="q95", combine=FALSE)
for (i in seq(plist)) { print(plist[[i]] + viridis::scale_color_viridis()) }
```

Since we are working with spatial data, you can visualize gene expression in the **spatial context**, showing where the genes are expressed on the tissue section:

```{r}
#| label: plotting_3

plist = ImageFeaturePlot(sc, features=features_to_plot, max.cutoff="q95", combine=FALSE)
for (i in seq(plist)) { print(plist[[i]] + viridis::scale_fill_viridis()) }
```

A **dot plot** allows for the visualization of gene expression across multiple clusters for several genes. We have already introduced the dot plot for exploring marker gene expression. To avoid redundancy, we will not revisit it in this section.

For a more detailed look, we can create a **heatmap** to visualize the expression of the same genes in a subset of cells (10000). This allows for a more focused analysis of gene expression patterns across specific clusters.

```{r}
#| label: plotting_5

set.seed(1)
cluster_cells = Cells(sc) %>% 
  sample(10000)

DoHeatmap(sc, features=features_to_plot, 
          cells=cluster_cells,
          group.colors=cal_pal50, 
          label=FALSE) + 
  scale_fill_viridis() +
  guides(colour="none")
```

```{r}
#| label: rm_1

# Remove since not used anymore and we need the memory.
rm(counts, data, scale_data, umap)

# R will automatically clear the memory of freed objects whenever it needs more space.
# But in our experience it is sometimes better to explicitly tell R to do this - especially when working with big objects.
gc()
```

::: {.callout-note}
## Assignment
Visualize cluster marker genes and known marker genes on the UMAP and in spatial context, and identify at least one cell type from it. 
:::

# Spatial clustering and domains

So far, we have explored the **fundamentals of analyzing spatial (and single-cell) transcriptomic data** using Seurat. The workflow introduced above serves is a good starting point.

However, a key limitation of this approach is that it treats each spatial cell independently, without considering its physical context within the tissue. Spatial analysis methods aim to address this by incorporating neighborhood information—enhancing biological interpretation and robustness. Several steps in our basic workflow can be extended to include this spatial context.

## What is BANKSY?

[BANKSY](https://github.com/prabhakarlab/Banksy) is a computational method specifically designed for spatial transcriptomics analysis. By integrating information from neighboring cells, BANKSY can reduce noise in gene expression data, distinguish between different cell types based on spatial context, and identify spatial domains. 

BANKSY represents each cell or cell not only by its own transcriptomic profile but also by the average expression of its local neighborhood. This dual representation captures both intrinsic cellular identity and microenvironmental context. The degree to which neighborhood information is incorporated is controlled by the lambda parameter:  

* Lower values (e.g., **lambda = 0.2**) emphasize immediate neighbors — useful for **cell typing**  
* Higher values (e.g., **lambda = 0.8**) incorporate broader spatial context — useful for **domain detection**  

Although BANKSY is a standalone tool with its own framework, it can be integrated into Seurat workflows using the `RunBanksy` function from the `SeuratWrappers` package. This allows users to apply BANKSY directly to Seurat objects.

In summary, BANKSY leverages the principle that a cell’s identity is shaped not only by its own gene expression but also by its surrounding environment. By adjusting a single hyperparameter, it provides flexibility to shift focus from cellular identity to higher-order tissue organization.

## Improve clustering with BANKSY

BANKSY can serve as an alternative to the default clustering approach we used earlier. In this step, we apply BANKSY with a low lambda value (0.2) to focus on local neighborhood information, and set the number of neighbors (k_geom) to 15.

**Note:** This step can be time- and memory-intensive.

```{r}
#| label: spatial_1

# remove.packages('Banksy')
# restart R
#remotes::install_github("prabhakarlab/Banksy")
#remotes::install_github("jleechung/seurat-wrappers")

sc = RunBanksy(sc,
               assay="Xenium", assay_name="banksy_cells",
               features="variable",
               lambda=0.2, k_geom=15,
               verbose=TRUE)

# Set it as default
DefaultAssay(sc) = "banksy_cells"

gc()
```

At this point, we have created a new assay `banksy_space`, containing the BANKSY-transformed data, and have set it as the default.

**⚠️ Important**: BANKSY fills the `scale.data` layer. Therefore do not call `ScaleData` on the BANKSY assay as this negates the effects of lambda. 

```{r}
#| label: spatial_2

sc
```

Next, we apply PCA, clustering, and UMAP using the BANKSY assay. We use the first 15 PCs and set a clustering resolution of 0.5, though these values can and should be adjusted depending on your dataset.

```{r}
#| label: spatial_3

# Save Xenium seurat_clusters
sc$xenium_seurat_clusters = sc$seurat_clusters

sc = RunPCA(sc, reduction.name="banksy_cells_pca", features=rownames(sc[["banksy_cells"]]), npcs=30, nfeatures.print=5)
sc = FindNeighbors(sc, reduction="banksy_cells_pca", dims=1:30)
sc = FindClusters(sc, resolution=0.5, algorithm="leiden")
sc = RunUMAP(sc, reduction="banksy_cells_pca", reduction.name="banksy_cells_umap", dims=1:30, return.model=TRUE)
sc$banksy_cells_clusters = sc$seurat_clusters

# BANKSY cluster colors
cols_banksy_cells_clusters = rev(cal_pal50)[1:length(levels(sc$banksy_cells_clusters))]
names(cols_banksy_cells_clusters) = levels(sc$banksy_cells_clusters)

# Reset
DefaultAssay(sc) = "Xenium"
```

Here we compare the results of default **original clustering versus BANKSY cell clustering**, first using UMAP plots:

```{r}
#| label: spatial_4_2

p1 = DimPlot(sc, reduction="umap", label=TRUE, group.by="xenium_seurat_clusters", cols=cols_clusters) + NoLegend() + ggtitle("Original clustering")
p2 = DimPlot(sc, reduction="banksy_cells_umap", label=TRUE, group.by="banksy_cells_clusters", cols=cols_banksy_cells_clusters) + NoLegend() + ggtitle("BANKSY cell clustering")
p1 | p2
```

And as spatial plots:

```{r}
#| label: spatial_5

p1 = ImageDimPlot(sc, group.by="xenium_seurat_clusters", cols=cols_clusters) + NoLegend() + ggtitle("Original clustering")
p2 = ImageDimPlot(sc, group.by="banksy_cells_clusters", cols=cols_banksy_cells_clusters) + NoLegend() + ggtitle("BANKSY cell clustering")
p1 | p2
```

::: {.callout-tip}
## Recommendations
BANKSY does not automatically produce better results. Always evaluate both clustering outputs in the context of your biological question and data quality.
:::

## Identifying spatial domains with BANKSY

In addition to improving clustering, BANKSY can be tuned to detect larger-scale tissue structures, often referred to as spatial domains. By increasing the `lambda` parameter and the number of neighbors (`k_geom`), BANKSY incorporates more spatial context and captures broader patterns of gene expression across the tissue.

In the following example, we set `lambda=0.8` and `k_geom=50` to include more neighboring cells in the analysis.

**Note:** We first remove the BANKSY assay from the previous run to conserve memory. This step is optional but will save us some memory resources.

```{r}
#| label: spatial_6

sc = RunBanksy(sc,
               assay="Xenium", assay_name="banksy_domains",
               features="variable",
               lambda=0.8, k_geom=50,
               verbose=TRUE)

# Set it as default
DefaultAssay(sc) = "banksy_domains"

gc()
```

As before, we perform PCA, clustering, and UMAP on the BANKSY assay. We use the first 10 PCs and a clustering resolution of 0.5, but these values should again be adjusted depending on your dataset:

```{r}
#| label: spatial_7

sc = RunPCA(sc, reduction.name="banksy_domains_pca", features=rownames(sc[["banksy_domains"]]), npcs=30, nfeatures.print=5)
sc = FindNeighbors(sc, reduction="banksy_domains_pca", dims=1:30)
sc = FindClusters(sc, resolution=0.5, algorithm="leiden")
sc = RunUMAP(sc, reduction="banksy_domains_pca", reduction.name="banksy_domains_umap", dims=1:30, return.model=TRUE)
sc$banksy_domains_clusters = sc$seurat_clusters

# BANKSY comain colors
cols_banksy_domains_clusters = rev(cal_pal50)[1:length(levels(sc$banksy_domains_clusters))]
names(cols_banksy_domains_clusters) = levels(sc$banksy_domains_clusters)

# Reset
DefaultAssay(sc) = "Xenium"
```

Here we compare the results of default **original clustering versus BANKSY domain clustering**, first using UMAP plots:

```{r}
#| label: spatial_4

p1 = DimPlot(sc, reduction="umap", label=TRUE, group.by="xenium_seurat_clusters", cols=cols_clusters) + NoLegend() + ggtitle("Original clustering")
p2 = DimPlot(sc, reduction="banksy_domains_umap", label=TRUE, group.by="banksy_domains_clusters", cols=cols_banksy_domains_clusters) + NoLegend() + ggtitle("BANKSY domain clustering")
p1 | p2
```

And as spatial plots:

```{r}
#| label: spatial_5_2

p1 = ImageDimPlot(sc, group.by="xenium_seurat_clusters", cols=cols_clusters) + NoLegend() + ggtitle("Original clustering")
p2 = ImageDimPlot(sc, group.by="banksy_domains_clusters", cols=cols_banksy_domains_clusters) + NoLegend() + ggtitle("BANKSY domain clustering")
p1 | p2
```

If your clustering results look similar to those from the previous BANKSY run (with `lambda=0.2`), this could indicate that the dominant tissue structures are already captured at a local level. We encourage you to systematically explore different parameter settings for `lambda`, `k_geom`, and `resolution` to assess their impact on clustering outcomes and spatial domain resolution.

## Can we evaluate the generated analyses?

We can try to evaluate the different approaches by visualizing known marker genes. Which of the produced results best reflects the different cell types we know? 

```{r}
#| label: spatial_8
#| fig.width: 10

# Which of the three clusterings best reflects our known biology?
plist = list()
plist[[1]] = DotPlot(sc, features=features_to_plot, group.by="xenium_seurat_clusters", cols=cols_clusters, scale=TRUE)
plist[[2]] = DotPlot(sc, features=features_to_plot, group.by="banksy_cells_clusters", cols=cols_banksy_cells_clusters, scale=TRUE)
plist[[3]] = DotPlot(sc, features=features_to_plot, group.by="banksy_domains_clusters", cols=cols_banksy_domains_clusters, scale=TRUE)
names(plist) = c("Original clustering", "BANKSY cell clustering", "BANKSY domain clustering")
print(features_to_plot)
for (i in seq(plist)) { 
  p = plist[[i]] + viridis::scale_color_viridis() + ggtitle(names(plist)[i]) + 
    theme(strip.text.x=element_blank()) 
  print(p)
}
```

```{r}
#| label: spatial_9

# Which of the three UMAPs visualizes features_to_plot best?
message("Original clusters")
plist = FeaturePlot(sc, features=features_to_plot, reduction="umap", ncol=2, max.cutoff="q95", combine=FALSE)
for (i in seq(plist)) { print(plist[[i]] & viridis::scale_color_viridis()) }

message("BANKSY cell clusters")
plist = FeaturePlot(sc, features=features_to_plot, reduction="banksy_cells_umap", ncol=2, max.cutoff="q95", combine=FALSE)
for (i in seq(plist)) { print(plist[[i]] & viridis::scale_color_viridis()) }

message("BANKSY domain clusters")
plist = FeaturePlot(sc, features=features_to_plot, reduction="banksy_domains_umap", ncol=2, max.cutoff="q95", combine=FALSE)
for (i in seq(plist)) { print(plist[[i]] & viridis::scale_color_viridis()) }

```

Since this BANKSY run is also complete, we again remove the assay and PCA reduction to free up memory:

```{r}
#| label: spatial_cleanup

# Remove BANKSY assay and pca since not used anymore and we need the memory.
DefaultAssay(sc) = "Xenium"
sc[["banksy_cells"]] = NULL
sc[["banksy_domains"]] = NULL
gc()
```

::: {.callout-tip}
## Recommended reading

If you want to learn more about BANKSY, we find these pages helpful: 

* https://github.com/satijalab/seurat-wrappers/blob/master/docs/banksy.md  
* https://github.com/prabhakarlab/Banksy?tab=readme-ov-file  

:::

# Cell type annotation by reference mapping

TODO: RAJINDER 

# Export results to Xenium Explorer

TODO: ANDREAS

# Session info

To enhance reproducibility and facilitate the sharing of your analysis, it is good practice to include information about the R session and the packages used:

```{r}
#| label: session_info

sessioninfo::session_info()
```

# Useful resources

- Best practices:
  - [Single-cell best practices](https://www.sc-best-practices.org/preamble.html)
  - [Orchestrating Single-Cell Analysis with Bioconductor](https://bioconductor.org/books/release/OSCA/)
  - [Orchestrating Spatial Transcriptomics Analysis with Bioconductor](https://lmweber.org/OSTA/)
- Seurat:
  - [Essential commands](https://satijalab.org/seurat/articles/essential_commands)
  - [Basic Single-cell Analysis Vignette](https://satijalab.org/seurat/articles/pbmc3k_tutorial)
  - [Seurat Xenium Analysis Vignette](https://satijalab.org/seurat/articles/seurat5_spatial_vignette_2)
- Tools:
  - [Xenium Explorer](https://www.10xgenomics.com/support/software/xenium-explorer/latest)
  - [SpaNorm](https://www.bioconductor.org/packages/release/bioc/html/SpaNorm.html)
  - [SpatialLeiden](https://spatialleiden.readthedocs.io/stable/index.html)
  - [Banksy](https://github.com/prabhakarlab/Banksy)


