---
title: "🧬 Spatial Transcriptomics Data Analysis"
format: html
---

# Welcome 

Welcome to our training on spatial transcriptomics data analysis!  

Spatial transcriptomics allows us to measure gene expression across tissue sections while preserving spatial information, helping us understand not just **what** genes are active, but **where** they are active.

In this practical session, we will explore how to analyze spatial transcriptomics data using **R** and **RStudio**. Whether you are new to spatial data or already familiar with single-cell workflows, this guide will walk you through the essential steps.

Feel free to use your personal workbook to make notes, track your progress, and revisit it after this workshop.

---

In this **first part**, we will explore the dataset, learn how to load it into R, and examine how the data is structured within a Seurat object. We will also perform some initial visualizations to familiarize ourselves with the dataset and its spatial features.

## Setting up your R environment

Before diving into the data, we need to load a few important R packages. These libraries will provide the tools we need for analysis and visualization.

```{r}
#| label: setup
#| message: false

library(Seurat)           # Main package for single-cell and spatial transcriptomics analysis
library(SeuratWrappers)   # Add-ons for Seurat
library(patchwork)        # Helps combine multiple ggplot2 plots into one figure
library(tidyverse)        # Essential collection of R packages, including ggplot2, dplyr, magrittr
library(grid)             # For other plots
library(viridis)          # Our continuous colour palette

# Colors
cal_pal50 = c("#Fa1a8e", "#009B7D", "#ff9933", "#7083a4", "#ffce45", "#015e05", 
              "#fedf25", "#d2b48c", "#bb55e1", "#6ec5ab", "#5d57af", "#143341", 
              "#761445", "#d65b5a", "#94043a", "#e7a6cd", "#204519", "#87afd1", 
              "#9b9a9d", "#f95b70", "#83c874", "#808080", "#452b5b", "#ecb100", 
              "#f46124", "#525252", "#4c84a3", "#00bfff", "#01b4c6", "#174d79", 
              "#a6a0f2", "#76facc", "#8491be", "#a32a2f", "#1c8859", "#2cc012", 
              "#35782b", "#9c6755", "#3b3960", "#eeb0a1", "#3e1e13", "#0064c3", 
              "#d81e4a", "#74646c", "#f675da", "#ffce45", "#ec7014", "#e50000", 
              "#000000", "#a4527c", "#041859")   

# Set default colours for ggplot2
options(ggplot2.discrete.colour=cal_pal50)
options(ggplot2.discrete.fill=cal_pal50)
options(ggplot2.continuous.colour="viridis")
options(ggplot2.continuous.fill="viridis")
options(ggplot2.binned.colour="viridis")
options(ggplot2.binned.fill="viridis")

# Increase maximum allowed object size for running analyses in parallel
options(future.globals.maxSize = 1e10)

```

# Our dataset

In this section, we will introduce the dataset we are using for this training session.

## Dataset overview

We use a **mouse brain dataset** from [10x Genomics](https://www.10xgenomics.com/):

-   Source: Downloaded from [here](https://www.10xgenomics.com/datasets/xenium-prime-fresh-frozen-mouse-brain)
-   Sample: Male C57BL/6 mouse, 9 weeks old, non-diseased
-   Tissue: Fresh-frozen brain tissue & embedding in OCT
-   Gene Panel: [Xenium Prime 5K Mouse Pan Tissue and Pathways Panel](https://www.10xgenomics.com/support/software/xenium-panel-designer/latest/tutorials/pre-designed-xenium-prime-5k)
-   Additional Application: [Cell Segmentation Staining](https://www.10xgenomics.com/support/in-situ-gene-expression/documentation/steps/assay/xenium-in-situ-gene-expression-with-morphology-based-cell-segmentation-staining-user-guide)
-   Platform: Xenium Analyzer
-   Processing: Xenium Onboard Analysis (v3.0.0.15)

The dataset is located in the `datasets/xenium_mouse_brain` directory within your project folder. If you need to download the dataset again, you can use the `datasets/download.R` script.

## Dataset files

In the folder `datasets/xenium_mouse_brain` you find data of the following categories: 

- raw data 
- processed data 
- additional (Xenium independent) data

### Raw data

These data come directly from the Xenium Analyzer and comprise metadata about the run, raw transcript data and images.

Metadata:

- `metrics_summary.csv`: QC report in CSV format
- `experiment.xenium`: experiment and run metadata (required for Xenium Analyzer)
- `gene_panel.json`: list of used panel probes and corresponding genes

Transcript data: 

- `transcript.zarr.gz`: raw transcript data with x, y, z coordinates for each transcript in compressed zarr format (efficient storage)
- `transcripts.parquet`: same content as above but in parquet format (faster input/output)

Images: 

- `morphology.ome.tif`: 3D image
- folder `morphology_focus/`: contains 2D autofocus images

### Processed data

The automatic Onboard Xenium Analysis processes the raw data and performs cell/ nuclei segmentation. The generated segmentation mask is used to assign each transcript based on its x, y, z coordinate to a specific cell. 
The results of this onboard analysis comprises the following folders and files: 

- `analysis_summary.html`: Quality Control (QC) report and initial analysis in HTML format 
- `cells.zarr.zip`: precise information of cell & nuclear segmentation (used for transcript assignment, in compressed zarr format) 
- `cell_boundaries.csv.gz` & `nucleus_boundaries.csv.gz`: information of cell & nuclear segmentation (used for visualization, coarser resolution) 
- `cells.tsv.gz`: cell metadata (for QC) 
- `cell_feature_matrix.*`: transcript counts per cell and gene in different formats

### Additional data

Additional data typically include microscopic images or additional measurements that were performed after the Xenium run.

This dataset contains: 

- `Xenium_Prime_Mouse_Brain_Coronal_FF_he_image.ome.tif`: H&E images of the Xenium slide 
- `Xenium_Prime_Mouse_Brain_Coronal_FF_he_imagealignment.csv`: alignment file

## Counts data

Let us take a closer look at the `cell_feature_matrix/` directory, which includes three important files:

- `barcodes.tsv.gz`: contains the names (or identifiers) for each cell in the dataset 
- `features.tsv.gz`: contains essential information about the genes, including their Ensembl IDs, symbols and gene types 
- `matrix.mtx.gz`: contains the actual counts data 
  - The first three rows are headers. The third row (=our first row) contains the total number of genes, cells and count entries (not the same as total number of counts!). 
  - All other rows contain the gene index (row in `features.tsv.gz`), cell index (row in `barcodes.tsv.gz`) and the count value. 
  - Only non-zero counts are stored.

We have a first look at the data to find out the number of identified cells, genes and measured counts:

```{r}
#| label: data_inspect

# inspect features.tsv.gz (=genes)
read.delim("datasets/xenium_mouse_brain/cell_feature_matrix/features.tsv.gz", 
           header=FALSE, nrows=6)

# get number of genes 
nrow(read.delim("datasets/xenium_mouse_brain/cell_feature_matrix/features.tsv.gz", 
                header=FALSE))

# inspect barcodes.tsv.gz (=cells)
read.delim("datasets/xenium_mouse_brain/cell_feature_matrix/barcodes.tsv.gz", 
           header=FALSE, nrows=5)

# get number of cells 
nrow(read.delim("datasets/xenium_mouse_brain/cell_feature_matrix/barcodes.tsv.gz", 
           header=FALSE))

# inspect matrix.mtx.gz (=counts)
read.delim("datasets/xenium_mouse_brain/cell_feature_matrix/matrix.mtx.gz", 
           skip=2, nrows=5, header=FALSE, sep=" ")

# get number of counts 
sum(read.delim("datasets/xenium_mouse_brain/cell_feature_matrix/matrix.mtx.gz", 
           skip=3, header=FALSE, sep=" ")$V3)

```

::: callout-tip
### Example

This dataset has 13,780 genes, 63,173 cells and 86,638,709 counts.\
Row 3: Gene #6 (Abca3) has two transcripts in cell #1 (barcode aaaalabd-1).
:::

::: callout-note
### Assignment

Any ideas why there are 5006 genes in the used gene panel, but 13,780 genes are given in the counts data?
:::

# The Seurat object

The R package [Seurat](https://satijalab.org/seurat/) is a widely used tool for single-cell and spatial transcriptomics analysis. It allows researchers to work with complex biological data and perform tasks like dimensionality reduction, clustering, and visualization.

Another popular option is the [scanpy](https://scanpy.readthedocs.io/en/stable/)/[squidpy](https://squidpy.readthedocs.io/en/stable/) package in Python.

## Importing the data into Seurat

The first step in analyzing your data is to import it into Seurat and transform it into a Seurat object (=`xenium.obj`):

```{r}
#| label: data_load_1

xenium.obj = Seurat::LoadXenium("datasets/xenium_mouse_brain",
                                fov="fov",
                                segmentations="cell", # can be "cell", "nucleus" or NONE
                                flip.xy=TRUE) # switch x & y to mirror the Xenium Explorer view

# rename the data
xenium.obj$orig.ident = "brain"
SeuratObject::Idents(xenium.obj) = "orig.ident"
```

Now, let us take a look at the Seurat object to understand its structure:

```{r}
#| label: data_load_2

xenium.obj
```

::: callout-note
### Questions

How many assays are in the dataset?\
How many layers are present in the "Xenium" assay?
:::

## What is stored in the Seurat object?

The Seurat object (=`xenium.obj`) organizes data into different components. These components facilitate spatial transcriptomics analysis and are crucial for data interpretation. 
The key components include: 

- **Cells and genes**: these represent the spatial units (cells) and the genes that are being measured 
- **Assay data**: contains the gene expression data (also possible: chromatin accessibility, surface proteins, etc.) 
- **Metadata for cells**: information about each cell (e.g., sample type, condition) 
- **Metadata for genes**: includes gene IDs, names, types, and descriptions 
- **Dimensionality reductions**: methods like PCA, UMAP, and tSNE that are used to reduce the complexity of the data 
- **Neighbors and graphs**: information about the relatedness between cells 
- **Images**: spatial coordinates for visualizing the tissue

## Cells and genes

We can retrieve the total number of genes, cells and counts stored in the Seurat object like this:

```{r}
#| label: data_load_3

# inspect genes
SeuratObject::Features(xenium.obj) %>% head(5)

# get number of genes
n_genes = nrow(xenium.obj)
n_genes

# inspect cells 
Seurat::Cells(xenium.obj) %>% head(5)

# get number of cells 
n_cells = ncol(xenium.obj)
n_cells

# get both information
dim(xenium.obj)

# get total number of counts
counts = SeuratObject::GetAssayData(xenium.obj, assay="Xenium", layer="counts")
n_counts = sum(counts)
n_counts

# check the counts for one specific gene and one specific cell
counts["Abca3", "aaaalabd-1"]

```

::: callout-tip
### Example

This dataset has 5,006 genes, 63,173 cells and 86,525,682 counts.\
Gene #6 (Abca3) has two transcripts in cell #1 (barcode aaaalabd-1).
:::

::: callout-tip
### Nomenclature

Note that sometimes the term **cells** is also referred to as barcodes, bins, nuclei.\
The term **gene** is also referred to as feature, protein, or peak, depending on the dataset.\
We will stick to cells and genes for this workshop.
:::

Furthermore, cells have a default identity which is used for plotting. It can be accessed with the `Idents` function. This is especially interesting when loading multiple datasets into one Seurat object. We have renamed this item when loading the Xenium data:

```{r}
#| label: data_load_4

SeuratObject::Idents(xenium.obj) %>% head(5)
```

## Assays

Assays store the actual data, for example gene expression counts. We can list the available assays in the Seurat object:

```{r}
#| label: data_assay_1

SeuratObject::Assays(xenium.obj)
```

To access a specific assay, we use the `GetAssay` function or the `[[]]` syntax:

```{r}
#| label: data_assay_2

Seurat::GetAssay(xenium.obj, assay="BlankCodeword")
xenium.obj[["BlankCodeword"]]
```

As before, we can also check the number of genes and cells in a specific assay like this:

```{r}
#| label: data_assay_3

assay = Seurat::GetAssay(xenium.obj, assay="BlankCodeword")
dim(assay)
```

The Seurat object has a default/active assay that is used for most functions. To check the active assay, we can use:

```{r}
#| label: data_assay_4

DefaultAssay(xenium.obj)
```

To change the active assay to another one, for example `"ControlCodeword"`, use:

```{r}
#| label: data_assay_5

DefaultAssay(xenium.obj) = "ControlCodeword"
DefaultAssay(xenium.obj)
```

For this workshop, we will focus on the assay `"Xenium"` for our analysis, because this assay contains the gene expression data:

```{r}
#| label: data_assay_6

SeuratObject::DefaultAssay(xenium.obj) = "Xenium"
SeuratObject::DefaultAssay(xenium.obj)
```

::: callout-tip
### Default assay

Most Seurat functions will automatically use the default assay, but you can always specify a different assay with the `assay=` argument.
:::

::: callout-note
### Questions

Which assays are present in our Seurat object?\
How many genes are present in each assay?
:::

## Xenium controls

The pre-designed 5K Xenium panel contains also pre-designed negative controls of different types: 

- The assay **BlankCodeword** contains codewords that are not used by any probe in the panel. These codewords are also labelled as "unassignedCodeword" (e.g. in the file `gene_panel.json`). n=8,096 
- The assay **ControlCodeword** comprises codewords that have not corresponding probes present in the panel, similar to blank codewords. These negative control codewords are used to calculate the negative control metrics and to adjust the genes' Q scores. Their presence (or ideally absence) determine the specificity of the decoding algorithm. n=609 
- The assay **ControlProbe** contains codewords with present probes in the probe pool. These probes are designed for non-biological targets and in a similar way as for "real genes" (\~3 probes per non-biological target). Non-biological targets should not be present in the measured tissue i.e. the probes should not be able to bind and give a signal. n=21 
- The assay **GenomicControl** comprises codewords again with present probes in the probe pool. These probes are designed to bind to intergenic genomic DNA. These regions are present in the measured tissue but not as transcripts and should be therefore not bound by the probes. n=40

The data of the ControlProbe assay and the GenomicControl assay indicate the specificity of the probe binding and potential off-target effects.

The different negative controls can be checked in the Xenium Explorer or the overview file `analysis_summary.html`. If everything looks good, we can remove them from our Seurat object to keep it smaller. The total control counts per cell are still present in the cell metadata of the assay "Xenium".

```{r}
#| label: data_control 

# remove different assays from a seurat object
xenium.obj[["BlankCodeword"]] = NULL
xenium.obj[["ControlCodeword"]] = NULL
xenium.obj[["ControlProbe"]] = NULL
xenium.obj[["GenomicControl"]] = NULL

```

::: callout-tip
### Note

Other gene panels (also custom gene panels) have different controls. But they are generally designed by 10x Genomics and provided to the user.
:::

## Assay data

The assay object stores data in different **layers**: 

- `counts`: Raw counts for each gene in each cell
- `data`: Normalized counts for better comparability across cells
- `scale.data`: Scaled data for better comparability across genes

We use the `GetAssayData` function to access the actual data:

```{r}
#| label: data_counts

counts = SeuratObject::GetAssayData(xenium.obj, assay="Xenium", layer="counts")
counts[1:10, 1:10]
```

::: callout-note
### Questions

What do you find special about this matrix?\
How many and which layers do we have in our assay object?
:::

## Cell metadata

Cell metadata store additional information about each cell, such as the sample or condition. You can access this metadata using the `[[]]` syntax:

```{r}
#| label: data_meta_1

xenium.obj[[]] %>% head(5)
```

For example, the `orig.ident` column stores the original identity of the cell, and the `nCount_Xenium` and `nFeature_Xenium` columns store the number of counts and genes for the cells for this assay.

You can access individual metadata columns like this:

```{r}
#| label: data_meta_2

xenium.obj$nCount_Xenium %>% head(3)
```

## Gene metadata

Gene metadata can be accessed using the `[[]]` syntax on the Seurat assay object:

```{r}
#| label: data_meta_3
#| message: true

assay = Seurat::GetAssay(xenium.obj, assay="Xenium")
assay[[]] %>% head(5)
```

In this example, no gene metadata is available for the assay. If available, you can access individual columns using the `$` syntax as we have done it for the cell metadata.

## Dimensionality reductions

Dimensionality reductions (e.g., PCA, UMAP, tSNE) are essential for visualizing high-dimensional data in a lower-dimensional space. You can list available reductions with the `Reductions` function:

```{r}
#| label: data_reduction

SeuratObject::Reductions(xenium.obj)
```

Currently, there are no reductions in the Seurat object. If they were present, you could access them using the `Reductions` function or the `[[]]` syntax.

## Neighbors and graphs

To identify how the cells are related (e.g., for clustering), Seurat computes neighbors and graphs. You can list available neighbors and graphs using:

```{r}
#| label: data_neighbors

SeuratObject::Neighbors(xenium.obj)
SeuratObject::Graphs(xenium.obj)
```

Currently, there are no computed neighbors or graphs. If available, you could access them with the `GetNeighbors` / `GetGraph` functions or the `[[]]` syntax.

## Images

For spatial data, images are stored in the Seurat object to visualize tissue sections and spatial coordinates. To list available images, use:

```{r}
#| label: data_image_1

SeuratObject::Images(xenium.obj)
```

By default, one image `fov` ("field of view") is present for the Xenium dataset. Additional images (such as H&E) can be loaded or defined by the `Crop` function which we will do later on. You can access a specific image using the `[[]]` syntax:

```{r}
#| label: data_image_2

xenium.obj[["fov"]]
```

When we refer to the default image in spatial transcriptomics, we usually mean the spatial coordinates associated with each cell. These coordinates tell us where each cell is located within the tissue section but not what the tissue looks like visually. The DAPI image itself it not automatically loaded into the Seurat object.

The image slot typically contains the following information: 

- cell localization (called `centroids`) 
- cell segmentation (called `segmentation`) 
- molecule coordinates 
- image metadata

Normally, Seurat also allows us to load and view the underlying histological images such as H&E images that were taken for the Xenium sample. It is very helpful to plot those morphology images along with gene expression patterns. The current Seurat version however does not support this functionality for Xenium data. You can view the combination of gene expression data and histological images in the Xenium Explorer.

## Plotting

Now that we have the data in Seurat, we can start visualizing gene expression using plots.

### Spatial plot for gene expression

We can also visualize gene expression in a spatial context using a **spatial feature plot**:

```{r}
#| label: data_plotting_1

# plot with default or viridis color scale 
p1 = Seurat::ImageFeaturePlot(xenium.obj, features = "Acox1", max.cutoff ="q95") + scale_y_reverse()
p2 = Seurat::ImageFeaturePlot(xenium.obj, features = "Acox1", max.cutoff ="q95") + scale_fill_viridis_c() + scale_y_reverse() 

p1 | p2 + patchwork::plot_layout(guides = "keep")

```

```{r}
#| label: data_plotting_2

# plot with default settings or in black and white
p1 = Seurat::ImageFeaturePlot(xenium.obj, features = "Acox1", max.cutoff ="q95") + scale_y_reverse()
p2 = Seurat::ImageFeaturePlot(xenium.obj, features = "Acox1", max.cutoff ="q95", cols=c("black","white")) + scale_y_reverse() 

p1 | p2 + patchwork::plot_layout(guides = "keep")

```

```{r}
#| label: data_plotting_3

# plot multiple genes at once 
Seurat::ImageFeaturePlot(xenium.obj, features = c("Aatf", "Abcb1a", "Acox1"), max.cutoff ="q95", cols=c("black","white"), combine=TRUE) & 
  scale_y_reverse()

```

::: callout-tip
### Plot size and resolution matter

You may observe visual artifacts in the image. This can occur when the image is plotted at a relatively low resolution (e.g. 150 dpi) and subsequently viewed on a high-resolution display. To address this issue, consider increasing the resolution using the Quarto `dpi` option. Alternatively, reducing the figure size may also help mitigate the problem.
:::

::: callout-note
### Questions

What happens without the function `scale_y_reverse()` and why do we add it to the plotting function?\
What are the advantages/ disadvantages of the different colors and when would you use which?
:::

### Violin plot for gene expression

We can create a **violin plot** of the same genes to show the spatial-independent overview of their expression:

```{r}
#| label: data_plotting_4

Seurat::VlnPlot(xenium.obj, features=c("Aatf", "Abcb1a", "Acox1"), layer="counts", pt.size=0) + Seurat::NoLegend()
```

::: callout-note
### Assignment

Find out the top 3 genes with the highest expression and create a **spatial feature plot** and a **violin plot** for them.
:::

::: callout-note
### (Bonus-)Assignment

Explore the structure of the Seurat object with the `str()` function.
:::

### Total counts

We can also plot the total number of counts (or other general QC stats) for the whole tissue:

```{r}
#| label: data_plotting_5

p1 = Seurat::VlnPlot(xenium.obj, features = c("nCount_Xenium"), layer="counts", pt.size = 0) + Seurat::NoLegend()
p2 = Seurat::ImageFeaturePlot(xenium.obj, features = c("nCount_Xenium"), max.cutoff ="q99") + scale_fill_viridis_c() + scale_y_reverse() 

p1 | p2 + patchwork::plot_layout(guides = "keep")

```

## Subset data

We can subset the Xenium data based on coordinates we can gain from the Xenium Explorer. Either we use the rectangular selection tool and get min and max values for the x and y coordinates or we use the freehand selection tool to get cell IDs and use them to subset the Seurat object.

The cell IDs are saved by the Xenium Explorer as csv file. We provide this file for this workshop. The Seurat object is subsetted to the provided cell IDs. All other data are excluded. It is therefore important to copy the Seurat object before subsetting to avoid data loss.

```{r}
#| label: data_subset_1

# load selected cell ID's
select_cells <- read.csv("datasets/xenium_mouse_brain/selections/freehand-selection_cellids.csv", skip = 2, header = TRUE)
head(select_cells, n=3)

# subset the seurat object to selected cell ID's 
# note: save as new object to not overwrite the full xenium object
xenium.obj.subset <- subset(xenium.obj, cells = select_cells$Cell.ID)

```

::: callout-note
### Question & Assignment

How many cells have been included in the selection?\
Plot the total counts for the selected cells.
:::

```{r}
#| label: data_subset_2

# hint 1: check chunk data_load_3 for getting number of cells
# hint 2: check chunk data_plotting_5 for plotting total counts

```

The alternative approach with a rectangular selection follows a different idea: A new image is defined with the provided x and y coordinates and can be used for plotting. The original count data is not altered.

The x and y coordinates can be defined and saved with the Xenium Explorer, as well. Alternatively, they can be "hard-coded" in R (which we will do in this workshop).

::: callout-tip
Note: Keep in mind that we have flipped the x and y axis when loading the data into the Seurat object to resemble the representation in the Xenium Explorer.
:::

```{r}
#| label: data_subset_4 

# define x and y coordinates
x_min = 4000
x_max = 6000
y_min = 200
y_max = 3100

crop1 <- SeuratObject::Crop(xenium.obj[["fov"]],
             x = c(x_min, x_max),
             y = c(y_min, y_max))

# save new image "crop1"
xenium.obj[["crop1"]] <- crop1

# check new image
Images(xenium.obj)

```

```{r}
#| label: data_subset_5

# plot the full and cropped image 
p1 = Seurat::ImageFeaturePlot(xenium.obj, fov = "fov", features = c("nCount_Xenium"), max.cutoff ="q99") + scale_fill_viridis_c() + scale_y_reverse() + ggtitle("full image")
p2 = Seurat::ImageFeaturePlot(xenium.obj, fov = "crop1", features = c("nCount_Xenium"), max.cutoff ="q99") + scale_fill_viridis_c() + scale_y_reverse() + ggtitle("crop1")

p1 | p2 
```

It might be helpful to plot not only features for whole cells but also cell boundaries and localization of single molecules, especially for very small cropped areas:

```{r}
#| label: data_subset_6

# define x and y coordinates
x_min = 4400
x_max = 4830
y_min = 1665
y_max = 1995

crop2 <- SeuratObject::Crop(xenium.obj[["fov"]],
             x = c(x_min, x_max),
             y = c(y_min, y_max))

# save new image "crop2"
xenium.obj[["crop2"]] <- crop2

# switch to cell boundaries (default: "centroids")
SeuratObject::DefaultBoundary(xenium.obj[["crop2"]]) <- "segmentation"

# plot total counts 
Seurat::ImageFeaturePlot(xenium.obj, fov = "crop2", features = c("nCount_Xenium"), max.cutoff ="q99") + scale_fill_viridis_c() + scale_y_reverse()

# plot all molecules of specific genes 
Seurat::ImageDimPlot(xenium.obj, fov = "crop2", axes = TRUE, border.color = "white", border.size = 0.1, cols = "polychrome",
    coord.fixed = FALSE, molecules = c("Prox1", "Gpc4", "Glis3")) + scale_y_reverse() 

```

# Session info

To enhance reproducibility and facilitate the sharing of your analysis, it is good practice to include information about the R session and the packages used:

```{r}
#| label: session_info

sessioninfo::session_info()
```

**Now that we have successfully loaded the data into Seurat and explored the structure of the Seurat object, it's time to move on to the next step: data analysis.**

**To free the occupied memory, please restart the R session in RStudio via Session → Restart R.**
